---
title: "Stats_Exp2"
output: html_document
date: '2024-03-22'
---

#Experiment 2: Release from Microcapsules
Let's get into my 2nd set of data
```{r}
library(readxl) #for loading Excel files
library(dplyr) #for data processing
library(here) #to set paths
library(tidyverse)
library(rstatix)
#path to data
data_location <- here::here("data","pilotwide.xlsx")
pilotwide <- read_excel(data_location)

#blanks vs. treatment
ggplot(pilotwide, aes(factor(day), supernatant)) + 
  geom_boxplot() + facet_wrap(~strain)

#separate treatment groups from blanks
g7 <- pilotwide %>% filter(strain == "G7")
g7<- g7 %>% rename(reactor=reactor_no)

#boxplots over time for supernatant
ggplot(g7, aes(factor(day), supernatant)) + 
  geom_boxplot() + facet_wrap(~coating)

#boxplots over time for microcapsule
capsule <- g7 %>% filter(treatment == "capsule")
ggplot(capsule, aes(factor(day), microcapsule)) + 
  geom_boxplot() + facet_wrap(~coating)

#supernatant stats
group_by(g7, coating) %>% 
  summarise(mean = mean(supernatant, na.rm = TRUE),
                                    sd = sd(supernatant, na.rm = TRUE))
#microcapsule stats
group_by(capsule, coating) %>% 
  summarise(mean = mean(microcapsule, na.rm = TRUE),
            sd = sd(microcapsule, na.rm = TRUE))
#these stats are misleading because they average across days of the entire experiment, so of course the sd is high. 
#need to see it broken down by day
#supernatant stats
g7 %>% group_by(coating, day) %>% 
  summarise(mean = mean(supernatant, na.rm = TRUE),
                                    sd = sd(supernatant, na.rm = TRUE))
#microcapsule stats
capsule %>% group_by(coating, day) %>% 
  summarise(mean = mean(microcapsule, na.rm = TRUE),
            sd = sd(microcapsule, na.rm = TRUE))
#let's visualize this with a graph
#for this, we need to make a dataframe with avg and sd
#by taking avg and sd, we assume this data is normal. Let's check that
#normality test
g7 %>% shapiro_test(supernatant) #overall, the data is NOT normal
hist(g7$supernatant) 
g7 %>% group_by(day) %>% shapiro_test(supernatant) #grouping by day and coating means sample size is less than 3 and test can't be run
#within days, data IS normal except for day 0
g7 %>% group_by(coating) %>% shapiro_test(supernatant) #normal for each group
#should we log transform to make data normal overall? 
g7_log<- g7
g7_log$suplog<- log10(g7_log$supernatant)
g7_log$microlog<- log10(g7_log$microcapsule)
hist(g7_log$suplog)
hist(g7_log$microlog)
shapiro.test(g7_log$suplog)
shapiro.test(g7_log$microlog)
#still not normal. 
```
With the small sample size of the experiment and the lack of normality, can I do parametric tests here? 
#visualizing
```{r}
data_location <- here::here("data","pilotsum.xlsx")
summ <- read_excel(data_location)

microcapsule<- summ %>% filter(type == "microcapsule") %>% filter(meas == "mScarlet")
supernatant<- summ %>% filter(type == "supernatant") %>% filter(meas == "mScarlet")
sample<- microcapsule %>% filter(strain == "G7")

#add grouped bar chart for supernatant
super<- supernatant %>% filter(strain == "G7")
ggplot(super, aes(fill=coating, y=avg_rfu, x = as.factor(day))) + 
  geom_bar(position="dodge", stat="identity")+
geom_errorbar(aes(ymin=avg_rfu-sd, ymax=avg_rfu+sd), width=.2,
              position=position_dodge(.9))+
  theme_classic() + 
  xlab("Day") + 
  ylab("RFU Value")+ 
  theme(axis.text.x=element_text(angle = 0, hjust = .5, size = 15),
        axis.title = element_text(size = 16),
        axis.text.y = element_text(size = 14),
        legend.title = element_blank(),
        legend.text = element_text(size = 14),
        plot.title = element_text(size = 18, hjust = 0.5),
        legend.position = "bottom", 
        panel.grid.minor.y = element_line(color = "grey", 
                                          linetype = "dashed")) +
  ggtitle("Bacterial Levels Released in Supernatant")+
  theme(strip.text = element_text(
    size = 14)) +
  scale_fill_manual(values = c("no coating" = "aquamarine4", 
                                "chitosan" = "olivedrab3", 
                                "free" = "coral1"))+
  scale_x_discrete(breaks=c(0,3,7,11,14,24,42))


#add grouped bar chart for within microcapsules
sample<- microcapsule %>% filter(strain == "G7")
ggplot(sample, aes(fill=coating, y=avg_rfu, x = as.factor(day))) + 
  geom_bar(position="dodge", stat="identity")+
  geom_errorbar(aes(ymin=avg_rfu-sd, ymax=avg_rfu+sd), width=.2,
                position=position_dodge(.9))+
  theme_classic() + 
  xlab("Day") + 
  ylab("RFU Value")+ 
  theme(axis.text.x=element_text(angle = 0, hjust = .5, size = 15),
        axis.title = element_text(size = 16),
        axis.text.y = element_text(size = 14),
        legend.title = element_blank(),
        legend.text = element_text(size = 14),
        plot.title = element_text(size = 18, hjust = 0.5),
        legend.position = "bottom", 
        panel.grid.minor.y = element_line(color = "grey", 
                                          linetype = "dashed")) +
  ggtitle("Bacterial Levels Within Microcapsules")+
  theme(strip.text = element_text(
    size = 14)) +
  scale_fill_manual(values = c("no coating" = "aquamarine4", 
                               "chitosan" = "olivedrab3", 
                               "free" = "coral1"))+
  scale_x_discrete(breaks=c(0,3,7,11,14,24,42))

library("writexl")
data_location4 <- here::here("data","super.xlsx")
write_xlsx(super, data_location4)

data_location5 <- here::here("data","micro.xlsx")
write_xlsx(sample, data_location5)
```

#ANOVAs
```{r}
g7_mod <- g7 %>% dplyr::select(day, coating, reactor, supernatant)
anova_test(data=g7_mod, dv=supernatant, wid=reactor, within="day", between="coating") 
#all effects significant

#need to look at effect of treatment!
g7_mod2 <- g7 %>% dplyr::select(day, coating, reactor, supernatant, treatment)
anova_test(data=g7_mod2, dv=supernatant, wid=reactor, within="day", between="treatment") 
#yes, significant effect of treatment, day, and treatment*day

cap_sup<- capsule %>% dplyr::select(day, coating, reactor, microcapsule, supernatant) %>% na.omit()
capsule_mod <- capsule %>% dplyr::select(day, coating, reactor, microcapsule)
anova_test(data=capsule_mod, dv=microcapsule, wid=reactor, within="day", between="coating") 
#sig effect of day on bacteria level in capsule (no effect of coating)

#does microcapsule concentration impact supernatant?
#anova_test(data=cap_sup, dv=supernatant, wid=reactor, within=c("day","microcapsule"), between="coating") #think this is not working because microcapsule is numerical and R can't "divide" into groups based on those values
#need a different statistical test to answer this question

#visualize microcapsule concentration vs supernatant concentration
ggplot(data = cap_sup, aes(x = microcapsule, y = supernatant, color = coating)) +
  geom_point() + theme_classic() + xlab("Microcapsule Concentration") + ylab("Supernatant Concentration") + 
  theme(panel.grid.minor.y = element_line(color = "grey", linetype = "dashed"))
```
Want to do some sort of regression to assess this relationship. But again, unsure about normality/linearity
```{r}
hist(cap_sup$supernatant) #does not look normal. use spearman rank correlation? 
shapiro_test(cap_sup$supernatant) #not normal
shapiro_test(cap_sup$microcapsule) #not normal
#NOT NORMAL
```
#nonporametric
```{r}
#Spearman's rank correlation
result2<- cor.test(cap_sup$microcapsule, cap_sup$supernatant, method = "spearman")
result2 
#strong relationship between concentration of microcapsule and concentration of supernatant

#day vs concentration, show capsule and supernatant
cap_sup_long<- gather(cap_sup, meas, value, microcapsule:supernatant, factor_key=TRUE)
ggplot(data = cap_sup_long) +
  geom_boxplot(aes(x=day, y=value, group=interaction(day, meas), fill=factor(meas))) + theme_classic() + xlab("Day") + ylab("Concentration") + 
  theme(panel.grid.minor.y = element_line(color = "grey", linetype = "dashed")) + facet_wrap(~coating)
```

# Diving deeper
```{r}
#Boxplots of g7 show that free vs capsule treatments have different initial values. What if we normalize to day 0 in order to compare? This would also account for difference due to using sodium citrate to dissolve microcapsules 
#look at super and sample, which have average and sd values
sup_n<- super 
chit<- sup_n %>% filter(coating == "chitosan")
alg<- sup_n %>% filter(coating == "no coating")
free<- sup_n %>% filter(coating == "free")

chit$initial <- 5.78125
alg$initial <- 24.54700
free$initial <- 7315.25000

all<- bind_rows(chit, alg, free)
all$ratio<- all$avg_rfu/all$initial
all$logratio<- log10(all$ratio)

#boxplots over time for supernatant
ggplot(all, aes(factor(day), ratio)) + 
  geom_boxplot() + facet_wrap(~coating)

ggplot(all, aes(factor(day), logratio)) + 
  geom_boxplot() + facet_wrap(~coating) #think this is the correct way to present the data

#let's try with microcapsule data
micro_n<- sample
chitm<- micro_n %>% filter(coating == "chitosan")
algm<- micro_n %>% filter(coating == "no coating")

chitm$initial <- 956.3750
algm$initial <- 1487.8750

allm<- bind_rows(chitm, algm)
allm$ratio<- allm$avg_rfu/allm$initial
allm$logratio<- log10(allm$ratio)

ggplot(allm, aes(factor(day), ratio)) + 
  geom_boxplot() + facet_wrap(~coating)

ggplot(allm, aes(factor(day), logratio)) + 
  geom_boxplot() + facet_wrap(~coating)

#saving data
library("writexl")
data_location4 <- here::here("data","all.xlsx")
write_xlsx(allm, data_location4)

data_location5 <- here::here("data","allm.xlsx")
write_xlsx(allm, data_location5)

```
This tells us that the increase in supernatant bacterial levels relative to day 0 is greater for the microcapsules rather than free bacteria (because free bacteria start with higher initial level) - what does presenting the data in this way mean? is it biologically relevant? it also shows that the capsules help maintain a higher relative level by day 42, compared to free (which we do care about). Basically, demonstrates that capsule is acting as a release vehicle, whereas free is instant delivery - IMPLICATION of day 42
- still need to rerun stats on ratio
# Percent release
```{r}
#using dataset g7 to perform manipulation to calculate ratio, since supernatant and microcapsule concentration are related
#manipulation only relevant for microcapsule groups, so let's filter
man<- g7 %>% mutate(ratio = supernatant/(supernatant+microcapsule))
#how does this look? 
ggplot(man, aes(factor(day), ratio)) + 
  geom_boxplot() + facet_wrap(~coating)

# ratio for free is automatically 1 (supernatant/supernatant + 0 since no microcapsule)
man <- man %>% mutate(ratio = replace_na(ratio, 1))

#plot again
ggplot(man, aes(factor(day), ratio)) + 
  geom_boxplot() + facet_wrap(~coating)
#need to log transform since these are ratios
man <- man %>% mutate(logratio = log10(ratio))

data_location4 <- here::here("data","release.xlsx")
write_xlsx(man, data_location4)

#plot logs
ggplot(man, aes(factor(day), logratio)) + 
  geom_boxplot() + facet_wrap(~coating)
```
# Redoing relative with duplicates
```{r}
#Boxplots of g7 show that free vs capsule treatments have different initial values. What if we normalize to day 0 in order to compare? This would also account for difference due to using sodium citrate to dissolve microcapsules 
#look at super and sample, which have average and sd values
sup_n<- g7 %>% select(!microcapsule)

test<- sup_n
glimpse(test)
test$reactor <- as.character((test$reactor))

test <- test %>% mutate(initial = case_when(endsWith(test$reactor, "1") ~ "5.5625",
                                            endsWith(test$reactor, "2") ~ "6.0000", 
                                            endsWith(test$reactor, "3") ~ "26.875", 
                                            endsWith(test$reactor, "4") ~ "22.219", 
                                            endsWith(test$reactor, "5") ~ "7599.35", 
                                            endsWith(test$reactor, "6") ~ "7031.15"))

relative<- test
glimpse(relative)
relative$initial <- as.numeric(relative$initial)

relative$ratio<- relative$supernatant/relative$initial
relative$logratio<- log10(relative$ratio)

#boxplots over time for supernatant
ggplot(relative, aes(factor(day), ratio)) + 
  geom_boxplot() + facet_wrap(~coating)

ggplot(relative, aes(factor(day), logratio)) + 
  geom_boxplot() + facet_wrap(~coating) #think this is the correct way to present the data

#let's try with microcapsule data
micro_n<- g7 %>% select(!supernatant) %>% filter(treatment == "capsule")

micro_n$reactor <- as.character(micro_n$reactor)
micro_n <- micro_n %>% mutate(initial = case_when(endsWith(micro_n$reactor, "1") ~ "866.410",
                                            endsWith(micro_n$reactor, "2") ~ "1046.340", 
                                            endsWith(micro_n$reactor, "3") ~ "1216.600", 
                                            endsWith(micro_n$reactor, "4") ~ "1759.150"))
relative_m<- micro_n
glimpse(relative_m)
relative_m$initial <- as.numeric(relative_m$initial)

relative_m$ratio<- relative_m$microcapsule/relative_m$initial
relative_m$logratio<- log10(relative_m$ratio)

#boxplots over time for supernatant
ggplot(relative_m, aes(factor(day), ratio)) + 
  geom_boxplot() + facet_wrap(~coating)

ggplot(relative_m, aes(factor(day), logratio)) + 
  geom_boxplot() + facet_wrap(~coating) #think this is the correct way to present the data


#saving data
library("writexl")
data_location4 <- here::here("data","relative_super.xlsx")
write_xlsx(relative, data_location4)

data_location5 <- here::here("data","relative_micro.xlsx")
write_xlsx(relative_m, data_location5)

```

